---
title: "Rmd Exercise Template"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The question you are answering

You use the `##` to create a header. Use the header to indicate the question you are answering. After the header, in this space here, annotate your answer, telling us what you did.

```{r chunk name}
## Load the data

#Load the `gapminder` dataset.


gapminder <- read.csv("~/Desktop/MSIA_bootcamp/bootcamp-2019/data/gapminder5.csv", stringsAsFactors = F)
```


## If Statement

Use an if() statement to print a suitable message reporting whether there are any records from 2002 in the gapminder dataset. Now do the same for 2012.

Hint: use the `any` function.

```{r if statement}
if(any(gapminder$year == 2002) == TRUE){
  print('There are records from 2002 present.')
}else{
  print('There are no records from 2002 present.')
}
if(any(gapminder$year == 2012) == TRUE){
  print('There are records from 2012 present.')
}else{
  print('There are no records from 2012 present.')
}

```


## Loop and If Statements

Write a script that finds the mean life expectancy by country for countries whose population is below the mean for the dataset

```{r loop and if 1}

overall_mean <- mean(gapminder$lifeExp)

for(i in unique(gapminder$country)){
  country_mean <- mean(gapminder$lifeExp[gapminder$country == i])
  
  if(country_mean<overall_mean){
    print(paste0("Mean Life Expectancy of Country ", i, " is ", round(country_mean, 4), "."))
  }
  
}



```

Write a script that loops through the `gapminder` data by continent and prints out whether the mean life expectancy is smaller than 50, between 50 and 70, or greater than 70.

```{r loop and if 2}


for(i in unique(gapminder$continent)){
  continent_mean <- mean(gapminder$lifeExp[gapminder$continent == i])
  
  if(continent_mean < 50){
    print(paste0('The mean Life Expectancy of continent ', i, ' -> ', continent_mean, ' is less than ', 50,'.' ))
  }else if(continent_mean <= 70){
    print(paste0('The mean Life Expectancy of continent ', i, ' -> ', continent_mean, ' is between ', 50,' and ', 70, '.' ))
  }else{
    print(paste0('The mean Life Expectancy pf continent ', i, ' -> ', continent_mean, ' is greater than ', 70,'.' ))
  }
  
}
```


## Exercise: Write Functions

Create a function that given a data frame will print the name of each column and the class of data it contains. Use the gapminder dataset.  Hint: Use `mode()` or `class()` to get the class of the data in each column. Remember that `names()` or `colnames()` returns the name of the columns in a dataset.

```{r function 1}
### Answer

info_func <- function(dataframe){
  for(column in colnames(dataframe)){
    print(paste0('Name: ', column, ' , Class: ', class(dataframe[,column])))
  }
}

info_func(gapminder)

```

Create a function that given a vector will print the mean and the standard deviation of a **vector**, it will optionally also print the median. Hint: include an argument that takes a boolean (`TRUE`/`FALSE`) operator and then include an `if` statement.

```{r function 2}

vector_stats <- function(vector, median=FALSE){
  print(paste0('Mean: ', mean(vector, na.rm = TRUE)))
  print(paste0('SD: ', sd(vector, na.rm = TRUE)))
  if(median==TRUE){
    print(paste0('Median: ', median(vector, na.rm = TRUE)))
  }
}

my_vec <- c(sample(seq(1,100), 20), rep(NA, 5))
vector_stats(my_vec, median=TRUE)

```

Use what you’ve learned so far to answer the following questions using the gapminder dataset. Be sure to include some visualizations!

1. What is the relationship between GDP per capita and life expectancy? Does this relationship change over time? (Hint: Use the natural log of both variables.)

```{r another plot1}
plot(gapminder$lifeExp ~ log(gapminder$gdpPercap),
     main = 'Relationship Between Life Expectancy and log GDP Per Capita',
     xlab = 'Log GDP Per Capita',
     ylab = 'Life Expectancy',
     col = 'blue')
```


From the scatter plot above, we can see a rough positive relationship between log GDP Per Capita and the associated Life Expectancy for a country/year pair. Concretely, as log GDP Per Capita for a country increases, so does the life expectancy of said country for a given year. 




2. Does the relationship between GDP per capita and life expectacy vary by continent? Make sure you divide the Americas into North and South America.








Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
