---
title: "Advanced R: Reshaping, merging, and data manipulation - exercises"
author: "Richard Paquin Morel and Ali Ehlen, adapted from exercises by Christina Maimone"
date: "`r Sys.Date()`"
output: html_document
params:
    answers: FALSE
---


```{r, echo=FALSE, eval=TRUE}
answers<-params$answers
```

```{r global_options, echo = FALSE, include = FALSE}
knitr::opts_chunk$set(echo=answers, eval=answers,
                      warning = FALSE, message = FALSE,
                      cache = FALSE, tidy = FALSE)
```

## Load the data

Read both California energy datasets. Make sure the `datetime` variable is in an appropriate data type (i.e. not character). 

```{asis}
### Answer 
```

```{asis}
**`dplyr`**
```

```{r}

library(dplyr)
library(lubridate)
library(readr)

generation <- read_csv('~/Desktop/MSIA_bootcamp/bootcamp-2019/data/ca_energy_generation.csv')
imports <- read_csv('~/Desktop/MSIA_bootcamp/bootcamp-2019/data/ca_energy_imports.csv')

generation2 <- fread('~/Desktop/MSIA_bootcamp/bootcamp-2019/data/ca_energy_generation.csv')

```


```{r}
library(dplyr)
library(lubridate)
generation_dp <- read.csv(here::here("data/ca_energy_generation.csv"), 
                          stringsAsFactors = F)
imports_dp <- read.csv(here::here("data/ca_energy_imports.csv"), 
                       stringsAsFactors = F)

generation_dp <- mutate(generation_dp, datetime = as_datetime(datetime))
imports_dp <- mutate(imports_dp, datetime = as_datetime(datetime))

```

```{asis}
**`data.table`**
```

```{r}
library(data.table)
library(lubridate)
#generation_dt <- fread(here::here("data/ca_energy_generation.csv"))
#imports_dt <- fread(here::here("data/ca_energy_imports.csv"))

#generation_dt[,datetime := as_datetime(datetime)]
#imports_dt[,datetime := as_datetime(datetime)]
```


## Merge and reshape the data

Merge the two datasets and then melt the resulting dataframe/datatable to make it tidy.

```{r}

library(reshape2)
long_gen <- melt(generation, id.vars = 'datetime',
                 variable.name = 'source',
                 value.name = 'usage')

head(long_gen)


```

```{r}

merged_energy <- merge(generation, imports, by = 'datetime')

long_merged_energy <- melt(merged_energy, 
                           id.vars = 'datetime',
                           variable.name = 'source',
                           value.name = 'usage')


```

```{r}
# We chose to use dplyr::inner_join here to illustrate how this can be done in a single piped operation, but you can also use base::merge
library(reshape2)
long_ca_energy_dp <- generation_dp %>%
  inner_join(imports_dp, by = "datetime") %>% 
  melt(id.vars = "datetime",
       variable.name = "source",
       value.name = "output")
```

```{asis}
**`data.table`**
```

```{r}

all_sources <- merge(generation_dt, imports_dt, by = "datetime")

long_ca_energy_dt <- melt(all_sources, 
                          id.vars = "datetime",
                          variable.name = "source",
                          value.name = "output")

# another option: advanced join
long_ca_energy_dt <- melt(generation_dt[imports_dt, on = "datetime"], 
                          id.vars = "datetime",
                          variable.name = "source",
                          value.name = "output")

```

## Creating new variables

Create a series of new variables: 

1. `day`, which is the year-month-day, without the hour. The `lubridate` function `as_date` will do this.
2. `log_output`, which is the natural log of the output.
3. **Challenge**: `per_output`, which is the percent of daily output represented by each observation. You will need to use `group_by` and to create a new variable with the total output for the day. (Make sure to use `ungroup()` after this!)

Bonus: If you are using `dplyr`, try to do this all in one pipe!

```{r}
temp <- merged_energy %>% 
  filter(large_hydro < 1000) %>% 
  mutate(log_coal = log(coal)) %>% 
  arrange(desc(coal))
head(temp)
```

```{r}

library(tidyverse)

merged_energy2 <- merged_energy %>% 
  mutate()

```

```{asis}
**`dplyr`**
```

```{r}
long_ca_energy_dp <- long_ca_energy_dp %>%
  mutate(day = as_date(datetime),
         log_output = log(output)) %>%
  group_by(day) %>%
  mutate(total_daily_output = sum(output, na.rm = T)) %>% 
  ungroup() %>% 
  mutate(per_output = output/total_daily_output)

# Check results
long_ca_energy_dp %>% select(day, log_output, per_output) %>% head()
```

```{r}

merged_energy2 <- merged_energy %>% 
  select(contains('hydro')) %>% 
  mutate(total_hydro = rowSums(., na.rm = T))
head(merged_energy2)

merged_energy3 <- merged_energy %>% 
  select(contains('hydro')) %>% 
  mutate(total_hydro = rowSums(., na.rm = T)) %>% 
  summarise(mean(total_hydro))
head(merged_energy3)


```

```{r}

merged_energy2 <- long_gen %>%
  filter(source %in% c('small_hydro', 'large_hydro', 'biogas', 'biomass')) %>% 
  group_by(source) %>% 
  summarise(mean(usage))
head(merged_energy2)
  



```

```{r}
long_ca_energy_dt[,day := as_date(datetime)]
long_ca_energy_dt[,log_output := log(output)]
long_ca_energy_dt[,per_output := output/sum(output, na.rm = TRUE), by = day]

# one command (stylistic/readability choice)
long_ca_energy_dt[,`:=`(day = as_date(datetime),
                        log_output = log(output), 
                        per_output = output/sum(output, na.rm = TRUE)), 
                  by = day]

```

## Summarizing and analyzing data

1. Which source has the greatest mean output by hour? (Hint: Use the `dplyr` verb `arrange(desc(variable))` to order the data frame so that the largest value of `variable` is first. Don't use `desc` and it arranges in ascending order. The `data.table` function is `setorder`.) Which has the least?
2. Which source has the greatest mean output by day? Which has the least? (Do not include zero values.)
3. Which sources has the greatest variance in usage over the course of a dataset? Which has the least? (Do not include zero values.)

```{asis}
### Answer
```

```{asis}
#### 1. Which source has the greatest mean output by hour? Which has the least?
```

```{asis}
**`dplyr`**
```

```{r}
long_ca_energy_dp %>%
  group_by(source) %>%
  summarize(mean_hourly = mean(output, na.rm = T)) %>%
  arrange(desc(mean_hourly))

long_ca_energy_dp %>%
  group_by(source) %>%
  summarize(mean_hourly = mean(output, na.rm = T)) %>%
  arrange(mean_hourly)
```

```{r}
library(data.table)
#head(generation[generation$natural_gas>5000 & generation$large_hydro > 2000,])

print(generation2[, total_hydro := small_hydro + large_hydro])
print(generation2[,.(mean(nuclear), mean(biogas))])
print(generation2[,median(solar), by = hour(datetime)])
print(generation2[solar>0, max(natural_gas), by = mday(datetime)])


```

```{r}
mean_hrly <- long_ca_energy_dt[,.(mean_hourly = mean(output)), by = source]
mean_hrly[mean_hourly == max(mean_hourly) | mean_hourly == min(mean_hourly)]

# another option
setorder(mean_hrly, -mean_hourly)
mean_hrly

# another option: chained together, no new variable
long_ca_energy_dt[,.(mean_hourly = mean(output)), by = source][mean_hourly %in% c(max(mean_hourly), min(mean_hourly))]

```

```{asis}
#### 2. Which source has the greatest mean output by day? Which has the least? (Do not include zero values.)
```

```{asis}
**`dplyr`**
```

```{r}
long_ca_energy_dp %>%
  filter(output>0) %>% 
  group_by(day, source) %>%
  summarize(mean_daily = mean(output, na.rm = T)) %>%
  arrange(desc(mean_daily))

long_ca_energy_dp %>%
  filter(output>0) %>% 
  group_by(day, source) %>%
  summarize(mean_daily = mean(output, na.rm = T)) %>%
  arrange(mean_daily)
```

```{asis}
**`data.table`**
```

```{r}

mean_dly <- long_ca_energy_dt[output > 0,.(mean_daily = mean(output)), by = .(source, day)]

mean_dly[mean_daily %in% c(max(mean_daily), min(mean_daily))]

```

```{asis}
#### 3. Which sources has the greatest variance in usage over the course of a dataset? Which has the least? (Do not include zero values.)
```

```{asis}
**`dplyr`**
```

```{r}
long_ca_energy_dp %>%
  filter(output>0) %>%
  group_by(source) %>%
  summarize(sd_output = sd(output, na.rm = T)) %>%
  arrange(desc(sd_output))

long_ca_energy_dp %>%
  filter(output>0) %>% 
  group_by(source) %>%
  summarize(sd_output = sd(output, na.rm = T)) %>%
  arrange(sd_output)
```

```{asis}
**`data.table`**
```

```{r}
var_by_source <- long_ca_energy_dt[output > 0,.(sd_output = sd(output)), by = source]

var_by_source[sd_output %in% c(max(sd_output), min(sd_output))]

```

```{asis}
There are also other ways to find min and max values:
```

```{r}

boxplot(output ~ source, data = long_ca_energy_dt,las=2)

```


## Analyzing renewable versus non-renewable energy sources

The dataset `regroup.csv` has information about which sources are considered renewable by the state of California. Use this dataset, along with yourdata manipulation skills, to explore the use of renewable and non-renewable sources. Annotate what your descisions for the analysis.

Hint: Use your merge skills to merge the CA energy data with the `regroup` data. Which variable should you join by?